Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-04-12 analysis/paper/paper.Rmd
crossing fingers, final edits before preparing for submission

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. Native Northern Pikeminnow have been estimated to consume \~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider their potential impacts on local Chinook Salmon populations. The Deadwater Slough is within a migration corridor for several Chinook Salmon populations as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough, an approximately 1.9 kilometer reach of the Salmon River, to be 19,499 in the fall and 10,352 in the spring, corresponding with the peak emigration windows of juvenile Chinook Salmon. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon. Assuming 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 61,409 juvenile Chinook Salmon, annually. After performing a sensitivity analysis, we estimated Northern Pikeminnow predation on juvenile Chinook Salmon was equivalent to 377 (95% CI: 161 - 935) returning adults. Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has a consequential impact on Chinook Salmon recovery in the Upper Salmon River.

@@

Predation on emigrating juveniles by piscivorous fishes is another important, and potentially under-estimated, source of salmonid mortality. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of recovery efforts [@Friesen1999; @Tabor1993; @Beamesderfer1996]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow-water habitats [@Wydoski2003] and thus have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Northern Pikeminnow are estimated to consume \~8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is within a section of the Salmon River containing historically important overwinter rearing habitat for juvenile Chinook Salmon *O. tshawytscha* and is part of the migratory pathway for upstream Chinook Salmon, the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015] and several populations of threatened Snake River steelhead. The slough is also inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*, that are potential sources of mortality for rearing and migrating salmonids.

@@

In this study, we estimated the abundance of Northern Pikeminnow in the Deadwater Slough and evaluated potential impacts to juvenile salmon emigrants, focusing on DSR and NRR Chinook Salmon. We hypothesized that high densities of piscivorous predators in the Deadwater Slough might explain the reduced survival observed for juvenile Chinook Salmon. To test this, our objectives were four-fold:

1.  Estimate the abundance of Northern Pikeminnow in the Deadwater Slough during the peaks of fall DSR and spring NRR juvenile emigrations;
2.  Document predation on juvenile Chinook Salmon during the DSR and NRR emigration periods using gastric lavage;
3.  Use an established bioenergetics approach to estimate consumption potential (grams) of the Northern Pikeminnow population on DSR and NRR Chinook Salmon emigrants at Deadwater Slough;
4.  Estimate how consumption of juvenile Chinook Salmon emigrants at Deadwater Slough by Northern Pikeminnow may impact adult returns to the Upper Salmon River.

@@

Multiple capture methods were employed during the fall 2019 survey to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. After evaluating all methods, angling was the most effective method for capturing Northern Pikeminnow while also minimizing potential impacts to ESA-listed adult steelhead that were present during our surveys. Less effective methods were abandoned in subsequent years and the following analyses will focus on fish captured by angling, unless otherwise noted.

@@

Surveys conducted in spring 2021 did not follow a mark-recapture study design so Northern Pikeminnow abundance was estimated as the ratio of mean total CPUE in the fall surveys to the total CPUE in the spring survey, then multiplied that ratio by the mean abundance from fall surveys.

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes occupying Deadwater Slough. Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks to be analyzed in a controlled environment. For each stomach sample, total wet weight (grams) was recorded, including all fish and non-fish items (e.g., macroinvertebrates, organic matter). Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones [@Hansel1988; @Frost2000] or were categorized as unknown. Approximately 5% of Northern Pikeminnow were euthanized for dissection (n = 75) after gastric lavage to validate the efficacy of the methodology. Additionally, contents of livewells were examined during fish processing to ensure food items were not regurgitated.

@@

Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules (J)/g [@Deslauriers2017]. Prey energy densities were fixed at 3,000 J/g for invertebrates and 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. Because we were unable to differentiate juvenile Chinook Salmon from other fish prey, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. The average TL of Northern Pikeminnow caught in Deadwater Slough during our study was converted to fork length (FL) and then to weight (grams) using a weight-length formula from @Parker1995.

@@

The proportion of fish in the Northern Pikeminnow's diet relative to non-fish prey items is unknown in Deadwater Slough. Therefore, we conducted a series of model runs with varying proportions of fish in the diet ranging from 30 - 90%, in 10% increments. These values were supported by observations elsewhere in the Columbia River basin (54-86%: @Shively1996; 48-86%: @Zimmerman1999; 37%: @Gray2001) that suggest fish are the majority of Northern Pikeminnow prey by volume. A model run was conducted for each combination of diet scenario and season, resulting in fourteen estimates of the total grams of fish consumed by an individual Northern Pikeminnow. To estimate the total biomass of fish consumed by Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated Northern Pikeminnow population sizes during fall and spring.

@@

We obtained an estimate of the number of juvenile Chinook Salmon consumed by dividing total biomass consumed by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon River. The average weights of DSR and NRR emigrants were calculated from fish captured at seven rotary screw traps upstream of Deadwater Slough during the fall and spring periods. Although the primary impact to Chinook Salmon in Deadwater Slough by Northern Pikeminnow is juvenile predation, the common metric to evaluate salmon recovery is adult returns. We quantified the potential impact of Northern Pikeminnow predation on adult returns by estimating the number of equivalent adults expected to return to Lower Granite Dam if predation by Pikeminnow was eliminated in the Deadwater Slough. To accomplish this, we multiplied the estimated total juvenile Chinook Salmon consumed by the median Granite-to-Granite smolt-to-adult return rate (SAR) of 0.00614 (SD = 0.00051) from @McCann2019 for Chinook Salmon in the Upper Salmon River.

@@

Using hook-and-line angling, we caught a total of `r prettyNum(sum(total_cpue$n_fish), big.mark = ",")` Northern Pikeminnow over the course of the study including `r sum(N_mods[[3]][[1]]$m) + sum(N_mods[[3]][[3]]$m)` recaptures; the overall CPUE was `r round(sum(cpue_df$n_fish) / (sum(cpue_df$total_effort_min) / 60), 2)` Northern Pikeminnow per angler hour (Table \@ref(tab:summary-tbl)). Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` in fall 2019 and from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` in fall 2020 (Table \@ref(tab:abund-tab)). We estimated larger populations of Northern Pikeminnow using the unadjusted multiple-census estimator compared to the single census and adjusted delayed-mixing multiple census estimators (Figure \@ref(fig:abund-fig)). Our sampling design most closely matched a multiple-census estimator; therefore, the Schnabel estimates were considered most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r prettyNum(round(mean(all_est$Lci[all_est$model == "Schnabel"])), big.mark = ",")` - `r prettyNum(round(mean(all_est$Uci[all_est$model == "Schnabel"])), big.mark = ",")`) using the unadjusted Schabel estimator. Using the delayed-mixing Schnabel estimator the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). All subsequent analyses use results from the adjusted delayed-mixing Schnabel estimator. For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

@@

We found that gastric lavage successfully removed food items in all dissected individuals, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` of those. Northern Pikeminnow captured during spring 2021 had a higher proportion of fish content in their diet compared to individuals captured in the fall surveys (Table \@ref(tab:gastric-tab)). The mean wet weight of total contents for an individual was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; SD = `r round(sd(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants made up `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of all stomach contents examined.

@@

During the fall DSR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish to maintain their body size, assuming 60% of their diet consisted of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet was fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Daily consumption rates were higher early in the fall and late in the spring and corresponded with water temperatures above 8 degrees Celsius (Figure \@ref(fig:bio-p)). Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` in the fall and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` in the spring 2021, and a diet consisting of 60% fish, the Northern Pikeminnow population was estimated to consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and `r round(tot_g_NRR, 1)` metric tons during the spring NRR emigration in Deadwater Slough.

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupies Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration even after accounting for the potential of delayed mixing of marked fish. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledged that their methods were unsuitable for sampling offshore, noting that water velocity, depth, and irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, the maximum depth at Deadwater Slough was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimated densities of Northern Pikeminnow suggest that slow-water reaches outside of reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available throughout the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the basin; additional information on contemporary piscine predator abundances may be needed [@Widener2021].

@@

```{r abund-data-lp, eval = F}
LP_mod %>%
  select(event_name, species, data) %>%
  distinct() %>%
  select(-species) %>%
  unnest(data) %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  rename(`Survey` = event_name) %>%
  kable(booktabs = T,
        format.args = list(big.mark = ","),
        caption = "Summary of Northern Pikeminnow angling survey data for the single-census mark-recapture estimator. M is the total number of fish caught in the first sample (week), marked and returned to the population; n is the total number of fish caught in the second sample (week), and m is the number of previously marked fish caught in the second week.") %>%
  kable_styling()

>>>>

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. Native Northern Pikeminnow have been estimated to consume \~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider their potential impacts on local Chinook Salmon populations. The Deadwater Slough is within a migration corridor for several Chinook Salmon populations as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough, an approximately 1.9 kilometer reach of the Salmon River, to be 19,499 in the fall and 10,352 in the spring, corresponding with the peak emigration windows of juvenile Chinook Salmon. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon. Assuming 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 61,409 juvenile Chinook Salmon, annually. After performing a sensitivity analysis, we estimated Northern Pikeminnow predation on juvenile Chinook Salmon was equivalent to 377 (95% CI: 161 - 935) returning adults. Given the relatively small size of the Salmon River, the high densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has a consequential impact on Chinook Salmon recovery in the Upper Salmon River.

@@

Predation on emigrating juveniles by piscivorous fishes is another important, and potentially under-estimated, source of salmonid mortality. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of recovery efforts [@Friesen1999; @Tabor1993; @Beamesderfer1996]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow-water habitats [@Wydoski2003] and thus have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Northern Pikeminnow are estimated to consume \~8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually [@Beamesderfer1996].

@@

Most predation studies have focused on reservoirs [@Murphy2021] and mainstem reaches [@Zimmerman1999; @Tabor1993; @Ward1995; @Shively1996] of the Columbia River basin whereas considerably less is known about the interaction between upstream, river-dwelling piscivorous fishes and their impacts on salmonid recovery above the reservoir systems [@Rubenson2020]. Upstream habitats containing slower water velocities and other attributes that support piscivorous predators may overlap with essential habitat for some salmonid species and life stages. One such example is Deadwater Slough, an approximately `r round(ddw_rkms, 1)` km long reach of unnaturally slow and deep water in the Salmon River, Idaho. The Deadwater Slough is within a section of the Salmon River containing historically important overwinter rearing habitat for juvenile Chinook Salmon *O. tshawytscha* and is part of the migratory pathway for upstream Chinook Salmon populations, the endangered Snake River Sockeye Salmon *O. nerka* population [@Axel2015], and several populations of threatened Snake River steelhead. The slough is also inhabited by piscivorous predators, including Northern Pikeminnow and Smallmouth Bass *Micropterus dolomieu*, that are potential sources of mortality for rearing and migrating salmonids.

@@

In this study, we estimated the abundance of Northern Pikeminnow in Deadwater Slough and evaluated potential impacts to juvenile salmon emigrants, focusing on DSR and NRR Chinook Salmon. We hypothesized that high densities of piscivorous predators in the Deadwater Slough might explain the reduced survival observed for juvenile Chinook Salmon. To test this, our objectives were four-fold:

1.  Estimate the abundance of Northern Pikeminnow in Deadwater Slough during the peaks of fall DSR and spring NRR juvenile emigrations;
2.  Document predation on juvenile Chinook Salmon during the DSR and NRR emigration periods using gastric lavage;
3.  Use an established bioenergetics approach to estimate consumption potential (grams) of the Northern Pikeminnow population on DSR and NRR Chinook Salmon emigrants at Deadwater Slough;
4.  Estimate how consumption of juvenile Chinook Salmon emigrants at Deadwater Slough by Northern Pikeminnow might impact adult returns to the Upper Salmon River.

@@

Multiple capture methods were employed during the fall 2019 survey to reduce selectivity and bias for species and size classes. Methods included raft electrofishing, fyke netting, snorkeling, and angling. After evaluating all methods, angling was the most effective method for capturing Northern Pikeminnow while also minimizing potential impacts to Endangered Species Act (ESA)-listed adult steelhead that were present during our surveys. Less effective methods were abandoned in subsequent years and the following analyses will focus on fish captured by angling, unless otherwise noted.

@@

Surveys conducted in spring 2021 did not follow a mark-recapture study design so Northern Pikeminnow abundance was estimated as the ratio of mean total CPUE in the fall surveys to the total CPUE in the spring survey, then multiplied that ratio by the mean abundance from fall surveys:

@@

Gastric lavage [@Foster1977] was used to examine the stomach contents of Northern Pikeminnow for the presence of juvenile Chinook Salmon and other fishes occupying Deadwater Slough. Immediately following lavage, stomach contents of individuals were preserved with 99% isopropyl alcohol in whirl-paks to be analyzed in a controlled environment. For each stomach sample, total wet weight (grams) was recorded, including all fish and non-fish items (e.g., macroinvertebrates, organic matter). Fish and fish remnants were identified to the lowest taxonomic unit using diagnostic bones [@Hansel1988; @Frost2000] or were categorized as unknown. Approximately 5% of Northern Pikeminnow were euthanized for dissection (n = 75) after gastric lavage to validate the efficacy of the methodology. Contents of livewells were examined during fish processing to ensure food items were not regurgitated.

@@

Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules (J)/g [@Deslauriers2017]. Prey energy densities were fixed at 3,000 J/g for invertebrates and 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. Because we were unable to differentiate juvenile Chinook Salmon from other fish prey in model runs, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. The average TL of Northern Pikeminnow caught in Deadwater Slough during our study was converted to fork length (FL) and then to weight (grams) using a weight-length formula from @Parker1995.

@@

The proportion of fish in the Northern Pikeminnow's diet relative to non-fish prey items is unknown in Deadwater Slough. Therefore, we conducted a series of model runs with varying proportions of fish in the diet ranging from 30 - 90%, in 10% increments. These values were supported by observations elsewhere in the Columbia River basin (54-86%: @Shively1996; 48-86%: @Zimmerman1999; 37%: @Gray2001) that suggest fish are the majority of Northern Pikeminnow prey by volume. A model run was conducted for each combination of diet scenario and season, resulting in fourteen estimates of the total grams of fish consumed by an individual Northern Pikeminnow. To estimate the total biomass of fish consumed by Northern Pikeminnow in Deadwater Slough, we multiplied the grams of fish consumed by an individual Northern Pikeminnow by the estimated Northern Pikeminnow population sizes during fall DSR and spring NRR emigrations.

@@

We obtained an estimate of the number of juvenile Chinook Salmon consumed by dividing total biomass consumed by the average weight of DSR (`r avg_DSR_g` g) and NRR (`r avg_NRR_g` g) emigrants in the Upper Salmon River. The average weights of DSR and NRR emigrants were calculated from fish captured at seven rotary screw traps upstream of Deadwater Slough during the fall and spring periods. Although the primary impact to Chinook Salmon in Deadwater Slough by Northern Pikeminnow is juvenile predation, the common metric to evaluate salmon recovery is adult returns. We quantified the potential impact of Northern Pikeminnow predation on adult returns by estimating the number of equivalent adults expected to return to Lower Granite Dam if predation by Pikeminnow was eliminated in Deadwater Slough. To accomplish this, we multiplied the estimated total juvenile Chinook Salmon consumed by the median Granite-to-Granite smolt-to-adult return rate (SAR) of 0.00614 (SD = 0.00051) from @McCann2019 for Chinook Salmon in the Upper Salmon River.

@@

Using hook-and-line angling, we caught a total of `r prettyNum(sum(total_cpue$n_fish), big.mark = ",")` Northern Pikeminnow over the course of the study including `r sum(N_mods[[3]][[1]]$m) + sum(N_mods[[3]][[3]]$m)` recaptures; the overall CPUE was `r round(sum(cpue_df$n_fish) / (sum(cpue_df$total_effort_min) / 60), 2)` Northern Pikeminnow per angler hour (Table \@ref(tab:summary-tbl)). Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` in fall 2019 and from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` in fall 2020 (Table \@ref(tab:abund-tab)). We estimated larger populations of Northern Pikeminnow using the unadjusted multiple census estimator compared to the single census and adjusted delayed-mixing multiple census estimators (Figure \@ref(fig:abund-fig)). Our sampling design most closely matched a multiple census estimator; therefore, the Schnabel estimates were considered most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r prettyNum(round(mean(all_est$Lci[all_est$model == "Schnabel"])), big.mark = ",")` - `r prettyNum(round(mean(all_est$Uci[all_est$model == "Schnabel"])), big.mark = ",")`) using the unadjusted Schabel estimator. Using the delayed-mixing Schnabel estimator the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). All subsequent analyses use results from the adjusted delayed-mixing Schnabel estimator. For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish per 100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

@@

We found that gastric lavage successfully removed food items in all dissected individuals, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")`<!--(`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%)--> stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` of those. Northern Pikeminnow captured during spring 2021 had a higher proportion of fish content in their diet compared to individuals captured in the fall surveys (Table \@ref(tab:gastric-tab)). The mean wet weight of total contents for an individual was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; SD = `r round(sd(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants made up `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of all stomach contents examined.

@@

During the fall DSR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish to maintain their body size, assuming 60% of their diet consisted of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet was fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Daily consumption rates were higher early in the fall and late in the spring and corresponded with higher water temperatures, especially above 8 degrees Celsius (Figure \@ref(fig:bio-p)). Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` in the fall and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` in the spring 2021, and a diet consisting of 60% fish, the Northern Pikeminnow population was estimated to consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and `r round(tot_g_NRR, 1)` metric tons during the spring NRR emigration in Deadwater Slough.

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupies Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration even after accounting for the potential of delayed mixing of marked and released fish. Those estimates translate to linear densities (fish per rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledged that their methods were unsuitable for sampling offshore, noting that water velocity, depth, and irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore, it is possible that their estimates may have been low. Conversely, the maximum depth at Deadwater Slough was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimated densities of Northern Pikeminnow suggest that slow-water reaches outside of reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available throughout the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the basin; additional information on contemporary piscine predator abundances may be needed [@Widener2021].

@@

```{r abund-data-lp, eval = F}
LP_mod %>%
  select(event_name, species, data) %>%
  distinct() %>%
  select(-species) %>%
  unnest(data) %>%
  mutate(across(event_name,
                ~ str_replace(., "_", " "))) %>%
  rename(`Survey` = event_name) %>%
  kable(booktabs = T,
        format.args = list(big.mark = ","),
        caption = "Summary of Northern Pikeminnow angling survey data for the single census mark-recapture estimator. M is the total number of fish caught in the first sample (week), marked and returned to the population; n is the total number of fish caught in the second sample (week), and m is the number of previously marked fish caught in the second week.") %>%
  kable_styling()
